---
title: "New Guinea as a laboratory for avian evolution"
author: Ethan B. Linck
date: |
  | Ornithology guest lecture
  | University of Wyoming Department of Zoology & Physiology
  | @ethanblinck
header-includes:
- \addtobeamertemplate{title page}{\centering \includegraphics[width=1\textwidth]{/Users/ethanlinck/Dropbox/misc_talks/images/banner_ngbirds.png}}{}
output: beamer_presentation
fontsize: 11pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## Weird Bird 

\begin{figure}
\includegraphics[width = \linewidth]{/Users/ethanlinck/Dropbox/misc_talks/images/satinbird.jpg}
\end{figure}

(Yellow-breasted satinbird, 625-2000m)  

## Uniqueness

"Physicists and chemists often have genuine difficulty in understanding the biologist's stress of the unique...individuality [is] so characteristic of the organic world, where all individuals are unique; all stages in the life cycle are unique; all populations are unique; all species and higher categories are unique; all interindividual contacts are unique; all natural associations of species are unique; and all evolutionary events are unique" (Mayr 1961, "Cause and Effect in Biology", *Science*)

# Why is New Guinea's avifauna unique, and why does it matter?

## Outline

- Overview of New Guinea's avian diversity 
- Geography
- Speciation
- Macroevolution 
- Conservation

# Overview of New Guinea's avian diversity 

# $Q_1$: How do we quantify diversity?

## Species and phylogenetic richness

\begin{figure}
\includegraphics[width = 0.5\linewidth]{/Users/ethanlinck/Dropbox/misc_talks/images/rabp.jpeg}
\end{figure}

- 943 species regionally (Gregory 2009)
- 789 from the mainland (>5% of global total)
- 2 (or more?) species undescribed
- 98 families
- 39 birds-of-paradise

(https://macaulaylibrary.org/asset/225784281)

## Endemism

\begin{figure}
\includegraphics[width = 0.5\linewidth]{/Users/ethanlinck/Dropbox/misc_talks/images/wapb.jpeg}
\end{figure}

- 406 mainland endemics (>50%)
- 7 endemic families (Melanocharitidae, Paramythiidae, Cnemophilidae, Rhagologidae Ifritidae, Eulacestomidae, Melampittidae)
- 75 endemic species in Bismarcks / Admiralties / Bougainville

(https://macaulaylibrary.org/asset/184660681)

## Ecological diversity

\begin{figure}
\includegraphics[width = 0.5\linewidth]{/Users/ethanlinck/Dropbox/misc_talks/images/noca.jpeg}
\end{figure}

- Cassowaries to sunbirds to harpy eagles
- Coastal mudflats to swamp forest to savannah to glacier line (>4500m)
- More frugivores than in the American tropics
- Elaborate plumage possibly a result of reduced pressure from carnivores

(https://macaulaylibrary.org/asset/90807631)

## Importance to evolutionary thought

\begin{figure}
\includegraphics[width = 0.25\linewidth]{/Users/ethanlinck/Dropbox/misc_talks/images/mayr.jpeg}
\end{figure}

- Ernst Mayr (Modern Synthesis: biological species concept, allopatric speciation)
- Jared Diamond (*Guns, Germs, and Steel*, but also work on community assembly rules, speciation, biogeography)
- E.O. Wilson (ants, but taxon cycle, Theory of Island Biogeography)

(https://doi.org/10.1038/nature03435)

# Geography & Geology

# $Q_2$: What aspects of geography and earth history are associated with diversity?

## Geography  

\begin{figure}  
\includegraphics[width = 0.9\linewidth]{/Users/ethanlinck/Dropbox/misc_talks/images/position.png}
\end{figure}  
(wikipedia)

## Geography  

\begin{figure}  
\includegraphics[width = 0.9\linewidth]{/Users/ethanlinck/Dropbox/misc_talks/images/newguinea_topo.png} 
\end{figure}  

(wikipedia)

## Geography  

- High mountains with east-west orientation  
- "Sky Islands"  
- Large lowland basins  
- Major river systems
- Shallow and deep seas  
- Off-shore islands (land-bridge and oceanic)  

## Tectonic History

\begin{figure}  
\includegraphics[width = 0.8\linewidth]{/Users/ethanlinck/Dropbox/misc_talks/images/tectonics.png}
\end{figure}  

(Baldwin, Fitzgerald, & Webb 2012)  

## Tectonic History

- Gondwanan isolation  
- Current position at geographic crossroads   
- Collision to form mountain ranges  
- Movement to warmer climates during last ice age  
- "[A]rguably one of the most tectonically complex regions of the world...its tectonic
evolution has global significance" (Baldwin, Fitzgerald, & Webb 2012) 

## 19 Major Ecoregions

\begin{figure}  
\includegraphics[width = 0.9\linewidth]{/Users/ethanlinck/Dropbox/misc_talks/images/ecoregions.jpeg}
\end{figure}  

([one earth](https://www.oneearth.org/bioregions/new-guinea-surrounding-islands-au13/))

# Speciation

# $Q_3$: How might geography influence speciation?

## Allopatric speciation in the highlands

\begin{figure}  
\includegraphics[width = \linewidth]{/Users/ethanlinck/Dropbox/misc_talks/images/pujolar.png}
\end{figure}  

(Pujolar et al. 2022)

## Parapatric speciation in the highlands

\begin{figure}  
\includegraphics[width = \linewidth]{/Users/ethanlinck/Dropbox/misc_talks/images/lincketal.png}
\end{figure}  

(Linck, Freeman, & Dumbacher 2020)

## Allopatric speciation in the lowlands

\begin{figure}  
\includegraphics[width = \linewidth]{/Users/ethanlinck/Dropbox/misc_talks/images/deiner.png}
\end{figure}  

(Denier et al. 2011)

## Diversification on islands

```{r, warning=FALSE, message=FALSE, echo=FALSE}
library(calibrate)
library(maptools)

area=c(262.5,329,37,1,0.46,0.53,0.1,0.001,0.06,0.11,0.92)
divergence=c(0.0327,0.0119,0.0158,0.0119,0.0034,0.0034,0.0062,0.0062,0.0034,0.0034,0.0052)
names=c("Rossel","Long","Nissan","Nauna","Hummock","Haszard","Eyon","Ginetu","Noapoi","Duchess","Panamoti")
islands=data.frame(area=area,divergence=divergence)
plot(log(area), log(divergence), pch=18, xlab="Island area (km^2)(log scale)", ylab="Genetic distance from nearest neighbor (log scale)", cex.main=1.5, cex.lab=1.25, main="Genetic distance from nearest neighbor \n as a function of home island area")
trend=lm(log(divergence) ~ log(area), data=islands)
pointLabel(log(area), log(divergence), labels = names, cex = 1,
           allowSmallOverlap = FALSE)
abline(trend)
modsum=summary(trend)
r2=modsum$adj.r.squared
my.p = modsum$coefficients[2,4]
mylabel = bquote(italic(R)^2 == .(format(r2, digits = 3)))
text(x = 19, y = 2.5, labels = mylabel)
rp = vector('expression',2)
rp[1] = substitute(expression(italic(R)^2 == MYVALUE), 
                   list(MYVALUE = format(r2,dig=3)))[2]
rp[2] = substitute(expression(italic(p) == MYOTHERVALUE), 
                   list(MYOTHERVALUE = format(my.p, digits = 2)))[2]
legend('topleft', legend = rp, bty = 'n')

# island height by divergence
# height=c(838,1304,45,129,15,10,20,1,30,40,20)
# divergence=c(0.0327,0.0119,0.0158,0.0119,0.0034,0.0034,0.0062,0.0062,0.0034,0.0034,0.0052)
# islands2=data.frame(height=height,divergence=divergence)
# plot(log(height), log(divergence), pch=18, xlab="Island height (m)(log scale)", ylab="Genetic distance from nearest neighbor (log scale)", cex.main=1, main="Genetic distance from nearest neighbor \n as a function of home island height")
# trend2=lm(log(divergence) ~ log(height), data=islands2)
# pointLabel(log(height), log(divergence), labels = names, cex = 0.7,
#           allowSmallOverlap = FALSE)
# abline(trend2)
# modsum2=summary(trend2) 
# r2.2=modsum2$adj.r.squared
# my.p.2=modsum2$coefficients[2,4]
# mylabel2 = bquote(italic(R)^2 == .(format(r2.2, digits = 3)))
# text(x = 19, y = 2.5, labels = mylabel2)
# rp2 = vector('expression',2)
# rp2[1] = substitute(expression(italic(R)^2 == MYVALUE), 
#                   list(MYVALUE = format(r2.2,dig=3)))[2]
# rp2[2] = substitute(expression(italic(p) == MYOTHERVALUE), 
#                   list(MYOTHERVALUE = format(my.p.2, digits = 2)))[2]
# legend('topleft', legend = rp2, bty = 'n')

# distance=c(185,360,540,360,1.5,1.5,27,27,80,80,85)
# divergence=c(0.0327,0.0119,0.0158,0.0119,0.0034,0.0034,0.0062,0.0062,0.0034,0.0034,0.0052)
# islands3=data.frame(distance=distance,divergence=divergence)
# plot(log(distance), log(divergence), pch=18, xlab="Distance from nearest neighbor (km)(log scale)", ylab="Genetic distance from nearest neighbor (log scale)", cex.main=1, main="Genetic distance from nearest neighbor \n as a function of distance between islands")
# trend3=lm(log(divergence) ~ log(distance), data=islands3)
# pointLabel(log(distance), log(divergence), labels = names, cex = 0.7,
#           allowSmallOverlap = FALSE)
#  abline(trend3)
# modsum3=summary(trend3) 
# r2.3=modsum3$adj.r.squared
# my.p.3=modsum3$coefficients[2,4]
# mylabel3 = bquote(italic(R)^2 == .(format(r2.3, digits = 3)))
# text(x = 19, y = 2.5, labels = mylabel3)
# rp3 = vector('expression',2)
# rp3[1] = substitute(expression(italic(R)^2 == MYVALUE), 
#                    list(MYVALUE = format(r2.3,dig=3)))[2]
# rp3[2] = substitute(expression(italic(p) == MYOTHERVALUE), 
#                    list(MYOTHERVALUE = format(my.p.3, digits = 2)))[2]
#  legend('topleft', legend = rp3, bty = 'n')
```

(Linck et al. 2016)

# Macroevolution

# $Q_4$: Which avifaunas are most similar to New Guinea?   

## Australasian bird diversity 

\begin{figure}  
\includegraphics[width = 0.8\linewidth]{/Users/ethanlinck/Dropbox/misc_talks/images/marki.png}
\end{figure}  

(Marki et al. 2017)

## The origin of songbirds

\begin{figure}  
\includegraphics[width = 0.8\linewidth]{/Users/ethanlinck/Dropbox/misc_talks/images/jonsson.png}
\end{figure}  

(Jønsson et al. 2011)

## Avian diversity in Oceania

\begin{figure}  
\includegraphics[width = 0.8\linewidth]{/Users/ethanlinck/Dropbox/misc_talks/images/andersen.png}
\end{figure}  

(Andersen et al. 2018)  

# Biodiversity Conservation 

## Logging and Oil Palm

\begin{figure}  
\includegraphics[width = 0.9\linewidth]{/Users/ethanlinck/Dropbox/misc_talks/images/logging.jpg}
\end{figure}  

## Climate Change 

\begin{figure}  
\includegraphics[width = 0.9\linewidth]{/Users/ethanlinck/Dropbox/misc_talks/images/biocultural.png}
\end{figure}  

(Cámara-Leret et al. 2019)

## Climate Change 

\begin{figure}  
\includegraphics[width = 0.9\linewidth]{/Users/ethanlinck/Dropbox/misc_talks/images/freemanfreeman.png}
\end{figure}  

(Freeman & Class Freeman 2014)

## Cultural significance

\begin{figure}  
\includegraphics[width = 0.8\linewidth]{/Users/ethanlinck/Dropbox/misc_talks/images/andersen.png}
\end{figure}  

## Hunting

\begin{figure}  
\includegraphics[width = 0.8\linewidth]{/Users/ethanlinck/Dropbox/misc_talks/images/nugl.png}
\end{figure} 

## 45,000+ Years of Habitation  

\begin{figure}  
\includegraphics[width = 0.9\linewidth]{/Users/ethanlinck/Dropbox/misc_talks/images/habitation.jpeg}
\end{figure}  

## Indigenous Land Tenure & Conservation  

\begin{figure}  
\includegraphics[width = 0.9\linewidth]{/Users/ethanlinck/Dropbox/misc_talks/images/pwm.jpeg}
\end{figure}  

## Why New Guinea birds?

- New Guinea's bird diversity is globally important
- It showcases the speciation process at small scales
- It has played a role in the global trajectory of avian evolution
- It has a complex, pressing conservation future

## More weird birds

\begin{figure}  
\includegraphics[width = 0.9\linewidth]{/Users/ethanlinck/Dropbox/misc_talks/images/robin.jpeg}
\end{figure}  
(garnet robin)

## More weird birds

\begin{figure}  
\includegraphics[width = 0.9\linewidth]{/Users/ethanlinck/Dropbox/misc_talks/images/dove.jpeg}
\end{figure}  
(mountain fruit dove)

## Thanks + resources

- People: Stanley Jacobs, Bulisa Iova, Georgia Kapui, Salape Tulai, Jack Dumbacher, Ben Freeman, Kate Schimel, Bruce Beehler, Lisa Dabek
- Places: [The New Guinea Binatang Research Center](http://baloun.entu.cas.cz/png/), University of Washington, California Academy of Sciences
- Slides & references: https://github.com/elinck/misc_talks/ 







